Frequency of exposure mediates the facilitation effect of cognateness in bilingual word acquisition

### 2023-02-01 manuscript/manuscript.qmd
Update manuscript

thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
fig-dpi: 1000
csl: "apa7.csl"
bibliography: "references.bib"
format:
  pdf:
    pdf-engine: lualatex
    number-sections: true
    whitespace: small
    linestretch: 1.5
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(conflicted)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(forcats)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(brms)
library(papaja)
library(scales)
library(stringdist)
library(marginaleffects)
library(lubridate)
library(bvqdev)

# load and process objects
# load targets as R objects
tar_load(participants)
tar_load(df)
tar_load(items)
tar_load(bvq_data)
tar_load(model_fit_4)
tar_load(model_log_liks)
tar_load(model_fit_4_prior)
tar_load(model_loos)
tar_load(rope_interval)
tar_load(posterior_description)

participants <- left_join(participants, select(bvq_data$logs, id, time, edu_parent))
items <- left_join(items, select(bvq_data$pool, item, language, semantic_category, class))

@@

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = ggplot2::scale_color_gradient,
        ggplot2.continuous.colour = ggplot2::scale_color_gradient)

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

@@

The foundations of word learning are in place early in age. Before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure; @bates1994developmental]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Despite the recent increase in language representation in the literature, most studies on early word acquisition have relied exclusively on data from monolingual children while neglecting, to a considerable extent, that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous findings pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close like Spanish and Catalan (both Romance languages) they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological similarity was measured by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the edit distance between two strings by counting the smallest number of insertions, deletions and substitutions one of the strings has to go through to become identical to the other [@levenshtein1966binary]. The resulting scores were then divided by the length of the longest string to bound the similarity scores between 0 and 1, and then entered this variable as a predictor as they modelled participants' vocabulary sizes. Among other findings, the authors reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14% cognateness), were able to produce more Dutch words than English-Mandarin bilinguals (1.97% cognateness) were able to produce in Mandarin.

@@

Floccia et al. pointed to *parallel activation* as a phenomenon that might potentially underpin their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Subsequent studies have also provided similar evidence in comprehension in children [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is is line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. Importantly, bilinguals seem to be more likely to acquire words for which their translation equivalent has been already learnt in the other language, suggesting that semantic similarity also facilitates word acquisition across languages [@bilson2015semantic].

If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity, but not phonological similarity. Evidence supporting and earlier age of acquisition for cognates is, to date, scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (remarkably larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

> "This  finding  also  provides  support  to  the  proposal  that  the  cognate advantage is due to cognates being acquired before non-cognates in early childhood (Costa et al., 2016), leading to an ease of processing later in life."

We identify two main reasons why an earlier age of acquisition for cognates than for non-cognates is an unwarranted conclusion from Floccia et al.'s results. First, the response variable used was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Second, all participants were aged ~24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue overcoming some of these pitfalls. Using a larger, longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster.

@@

@@

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strenghtened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*) does not undergo the such facilitation. Both word-forms are fairly disimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](hypotheses.jpg){#fig-hypotheses}

@@

n_item_language <- count(bvq_data$pool, language) %>%
    pull(n) %>%
    set_names(c("catalan", "spanish"))
```

@@

```{r}
#| label: participants
#| echo: false
#| message: false
#| warning: false
#| include: false
# load packages
n_participants <- df %>%
    distinct(id, age) %>%
    count(id) %>%
    pull(n) %>%
    table()
n_responses <- nrow(distinct(df, id, age))
dates <- as_date(range(participants$time_stamp, na.rm = TRUE))
```

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child were aged between `r floor(min(df$age))` and `r ceiling(max(df$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan [see @tbl-participants for a summary of the distribution of participants across ages and degrees of exposure to Catalan]. All families gave informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a small present.

```{r}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan. Information about DoE was provided by families before filling the questionnaire. A 100% indicated that the participant was exclusively exposed to Catalan, and never to Spanish. A 0% indicates that the participant was not exposed to Catalan ever, and rather most of the time to Spanish. A 50% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100%, 0%, and 50% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants %>%
    filter(id %in% df$id) %>% 
    mutate_at(vars(starts_with("doe_")), function(x) {
        floor(x * 10) / 10
    }) %>%
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%", "25-50%", "50-75%", "75-100%"),
            include.lowest = TRUE
        ) %>%
            as.character()
    ) %>%
    count(age, doe_catalan) %>%
    arrange(desc(doe_catalan)) %>%
    pivot_wider(names_from = age, values_from = n) %>%
    gt() %>%
    summary_rows(
        columns = 2:7,
        fns = list("Total" = ~ sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>%
    sub_missing(everything(), everything(), missing_text = "--") %>%
    cols_label(doe_catalan = "DoE Catalan") %>%
    tab_spanner("Age (months)", 2:8) %>%
    tab_footnote(
        "This proportion is complementary to the degree of exposure to Spanish, with the exception of those participants who were also exposed to a third language up to 10% of the time",
        cells_column_labels(columns = doe_catalan)
    ) %>%
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

@@

```{r}
#| label: sample-sizes
#| echo: false
#| message: false
#| warning: false
#| include: false
n_obs <- nrow(df)
sum_obs <- df %>%
    mutate(te = factor(te)) %>%
    count(te) %>%
    summarise(across(n, lst(median, min, max), .names = "{.fn}"))

n_items <- ifelse(grepl("cat_", distinct(df, item)$item),
                  "Catalan",
                  "Spanish") %>%
    table()

n_te <- length(unique(df$te))

>>>>

thanks: "This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances."
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
fig-dpi: 1000
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
format:
  pdf:
    pdf-engine: xelatex
    whitespace: small
    linestretch: 1.5
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages
library(ggplot2)
library(patchwork)
library(lubridate)
library(brms)
library(glue)
library(dplyr)
library(tidyr)
library(purrr)
library(gt)
library(stringr)
library(tibble)
library(tidybayes)
library(janitor)
library(scales)
library(bvqdev)

# load and process objects
# load targets as R objects
tar_load(participants)
tar_load(responses)
tar_load(items)
tar_load(bvq_data)
tar_load(model_fit_4)
tar_load(model_log_liks)
tar_load(model_fit_4_prior)
tar_load(model_loos)
tar_load(posterior_draws)
tar_load(marginal_effects_epreds)

items <- bvq_data$pool %>% 
    mutate(item = str_remove(item, "cat_|spa_")) %>% 
    select(item, language, semantic_category, class) %>% 
    right_join(items, by = c("item", "language"))

@@

options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient)

# resolve namespace conflicts
# conflict_prefer("filter", "dplyr")
# conflict_prefer("select", "dplyr")
```

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure; @bates1994developmental]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Despite the recent increase in the representativeness of the samples in the literature, most studies on early word acquisition have relied exclusively on data from monolingual children while neglecting, to a considerable extent, that a substantial proportion of the world population acquires more than one language since early ages [@grosjean2021extent]. Previous findings pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals do. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two typologically relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close like Spanish and Catalan (both Romance languages) they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological similarity was measured by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. The resulting scores are then divided by the length of the longest string to normalise the similarity scores between 0 and 1, and then entered this variable as a predictor as they modelled participants' vocabulary sizes. Among other findings, the authors reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14% cognateness), were able to produce more Dutch words than English-Mandarin bilinguals (1.97% cognateness) were able to produce in Mandarin.

@@

Floccia et al. pointed to *parallel activation* as a phenomenon that might potentially underpin their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

This account is especially plausible in the light of (1) the evidence pointing to bilingual toddlers' lexicon consisting of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022translation; @pearson1995crosslanguage], and that (2) there is experimental evidence of parallel activation during word comprehension in children and toddlers [e.g., @vonholzen2012language; @poulin-dubois2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is, to date, scarce nonetheless. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (remarkably larger) sample size. Their results suggest a cognateness effect on vocabulary size, but their study was not aimed at testing the effect of cognateness on age of acquisition at the word-level. Their response variable was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was no longer available. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue overcoming some of these pitfalls. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

## A mechanistic account for the cognate facilitation in word acquisition

In this study, we suggest a mechanistic account for the cognate facilitation effect. We build this account from the assumption that bilinguals use fundamentally similar mechanisms than monolinguals to learn words. In both cases, learning a word requires at least two events to take place. First, a representation of the word-form, encoding its segmental information to sufficient detail, must be stored.

This can be accomplished via bottom-up mechanism since early ages. Even if words are embedded in a continuous stream of speech sounds, in which there are no clear boundaries between words, infants can segment discrete word-forms. For instance, they may rely on their familiarity with the typical stress pattern [@jusczyk1995infants], the phonotactic constraints [@friederici1993phonotactic], or the distributional information of their native language [@saffran1996statistical] to detect word boundaries. A second event for word learning to take place is the association between the familiar word form and its referent. This task is far from trivial for many reasons. Referents are rarely labelled ostensively: children are exposed to most instances of a word in referentially ambiguous contexts or in the absence of their referent. Many words are also presented along remarkably variable exemplars of their referent, and even refer to abstract referents such as routines or play interactions (e.g., *uh-oh*; @casey2021moving). Infants nonetheless need few learning instances to create new word-referent associations. The extent to which they do so relying on bottom-up mechanisms such as cross-situational learning [@yu2007rapid], or on constraints and biases that guide their hypotheses about how words and referents are related [@golinkoff1994early], remains a matter of debate. Either way, there is evidence of word-referent links as early as six months of age [@bergelson2012months; @bergelson2015early]. From a connectionist perspective, a word is said to be acquired when the association between its word-form and its conceptualised referent (i.e. semantic representation) has reached some level of associative strength. As infants accumulate experience with word-referent pairings, the association strength between both increases, and this leads to observable events such as infants' showing a stronger preference for a named object compared to unnamed competitors, or their parents being more likely to report such word as acquired in a vocabulary checklist.

Bilinguals deal with a different linguistic landscape than monolinguals, but arguably use similar mechanisms and do so under common constraints [@curtin2011bilingual]. Bilingual word acquisition entails similar events as described above, with the exception that many referents are

they seem to reach their developmental goals of word learning at similar ages than monolinguals. Bilinguals store words

@@

We argue that the cognateness facilitation effect takes place

@@

![Schematic representation of the hypothesised cognate facilitation effect on word acquisition. To the left, the acquisition of a cognate translation equivalent (*flower*-*flor*) is affected by the phonological similarity between both word forms. When the child hears *flower*, activation spreads across languages to activate *flor* too. As a result, after exposure to *flower* not only the association between *flower* and its referent is strenghtened, but also that of *flower* and its common referent. To the right, the acquisition of a non-cognate translation equivalent (*cow*-*vaca*) does not undergo the such facilitation. Both word-forms are fairly disimilar, so that when the child hears *cow*, the activation of *vaca* is unlikely, rendering the effect of phonological similarity negligible.](img/hypotheses.jpg){#fig-hypotheses}

@@

n_item_language <- bvq_data$pool %>% 
    count(language) |>
    pull(n) |>
    set_names(c("catalan", "spanish"))
```

@@

```{r}
#| label: participants
#| echo: false
#| message: false
#| warning: false
#| include: false
# load packages
n_participants <- responses |>
    distinct(id, age) |>
    count(id) |>
    pull(n) |>
    table()

n_responses <- nrow(distinct(responses, id, age))

dates <- participants$time_stamp %>% 
    range(na.rm = TRUE) %>% 
    as_date()
```

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child were aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan [see @tbl-participants for a summary of the distribution of participants across ages and degrees of exposure to Catalan]. All families gave informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a small present.

```{r}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan. Information about DoE was provided by families before filling the questionnaire. A 100% indicated that the participant was exclusively exposed to Catalan, and never to Spanish. A 0% indicates that the participant was not exposed to Catalan ever, and rather most of the time to Spanish. A 50% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100%, 0%, and 50% would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."
participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) %>% 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, values_from = n) |>
    gt() |>
    summary_rows(
        columns = 2:7,
        fns = list("Total" = ~ sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) |>
    sub_missing(everything(), everything(), missing_text = "--") |>
    cols_label(doe_catalan = "DoE Catalan") |>
    tab_spanner("Age (months)", 2:8) |>
    tab_footnote(
        "This proportion is complementary to the degree of exposure to Spanish, with the exception of those participants who were also exposed to a third language up to 10% of the time",
        cells_column_labels(columns = doe_catalan)
    ) |>
    tab_style(cell_text(align = "left"),
              list(cells_body(doe_catalan),
                   cells_column_labels(doe_catalan)))
```

@@

```{r}
#| label: sample-sizes
#| echo: false
#| message: false
#| warning: false
#| include: false
n_obs <- nrow(responses)

sum_obs <- responses |>
    mutate(te = factor(te)) |>
    count(te) |>
    summarise(across(n, lst(median, min, max), .names = "{.fn}"))

n_items <- distinct(responses, language, item) |> 
    count(language) |> 
    group_split(language) |> 
    set_names(c("Catalan", "Spanish")) |>
    map_int("n")

n_te <- length(unique(responses$te))
